---
title: "annotate DEGlist with 2015 Phel blastx and GO"
output: html_document
date: "2022-11-01"
---
Rmd to annotate the DEGlist made by comparing Healthy (n=8) v Sick (n=8) 2021 coelomocyte RNA libraries to and the 2015 Phel transcriptome for counts

2015 Phel transcriptome blast output annotated with GOslim in jupyter notebook: https://github.com/grace-ac/project_pycno/blob/main/code/06-DEGlist-healthy_v_sick_annot_2015Phel_transcriptome.ipynb

The annotated 2015 Phel trancripomte with BLASTx and GO: https://raw.githubusercontent.com/grace-ac/project_pycno/main/analyses/DESeq2/2015Phel/Blastquery-GOslim.tab?token=GHSAT0AAAAAABZLW7F4A23IFERNKBSBZTFWY3AREDA

```{r}
library(tibble)
```

In this Rmd I'll:
`join` the two files based on the Phel contig 

```{r}
degs <- read.delim("../analyses/DESeq2/2015Phel/DEGlist_healthy-vs-sick.tab", sep = '\t')
head(degs)
```
Make the rownames into a column called "Phel_contig"
```{r}
degs <- tibble::rownames_to_column(degs, "Phel_contig")
head(degs)
```

Read in Blastquery-GOslim.tab of Phel transcriptome from Up in Arms:
```{r}
phelannot <- read.delim("../analyses/DESeq2/2015Phel/Blastquery-GOslim.tab", sep = '\t', header = FALSE)
head(phelannot)
```

Rename columns: 
```{r}
cols <- c("Phel_contig", "GO_ID", "GO_slim", "CC_MF_BP")
colnames(phelannot) <- cols
head(phelannot)
```

`join` degs and phelannot by "Phel_contig" column
```{r}
degsannot <- left_join(degs, phelannot, by = "Phel_contig")
head(degsannot)
```

32367 rows and 10 columns... 

Write out table:
```{r}
write.table(degsannot, file = "../analyses/DESeq2/2015Phel/DEGlist_annot_4x4_healthyvsick.tab", row.names = FALSE, sep = '\t', quote = FALSE)
```


20221205
NEED TO DO THIS WITH BIG TABLE
Read in .csv file that was made in excel by openeing DEGlist_annot.tab and filtering out only contigs with biological process GOids and sorting from lowest to highest log2fold change
```{r}
#degsannotBP <- read.delim("../analyses/DESeq2/2015Phel/DEGlist_annot_onlyBP.csv", sep = ',')
head(degsannotBP)
```

Subset into 2 tables --> one with all the negative log2foldchange values and one with all the positive log2foldchange values      
negative will be ones that are less expressed in the Healthy samples compared to the sick samples           
positive will be contigs that are more expressed in the Healthy samples compared to the sick samples     

```{r}
attach(degsannotBP)
degsLESShealthy <- degsannotBP[ which(log2FoldChange < 0),]
detach(degsannotBP)
head(degsLESShealthy)
```
2277 rows. Some Phel_contigs are repeats because multiple BP GOIDs per Phel_contig


```{r}
attach(degsannotBP)
degsMOREhealthy <- degsannotBP[ which(log2FoldChange > 0),]
detach(degsannotBP)
head(degsMOREhealthy)
```

5714 rows. Some Phel_contigs are repeats because multiple BP GOIDs per Phel_contig

Write out both tables:

```{r}
write.table(degsLESShealthy, file = "../analyses/DESeq2/2015Phel/DEGs_LESS-healthy-annot.tab", sep = '\t', quote = FALSE, row.names = FALSE)
```

```{r}
write.table(degsMOREhealthy, file = "../analyses/DESeq2/2015Phel/DEGs_MORE-healthy-annot.tab", sep = '\t', quote = FALSE, row.names = FALSE)
```

20221204
annotate 120 degs from comparing the 4 tank exposed libraries to the 4 experiment 2 inoculate live unfiltered exposure:   

read in list of degs:
```{r}
degstvs <- read.delim("../analyses/DESeq2/2015Phel/DEGlist_tankexposed_v_sick_4x4_pval.tab", sep = '\t')
head(degstvs)
```

Make the rownames into a column called "Phel_contig"
```{r}
degstvs <- tibble::rownames_to_column(degstvs, "Phel_contig")
head(degstvs)
```

Read in Blastquery-GOslim.tab of Phel transcriptome from Up in Arms:
```{r}
phelannot <- read.delim("../analyses/DESeq2/2015Phel/Blastquery-GOslim.tab", sep = '\t', header = FALSE)
head(phelannot)
```
Rename columns: 
```{r}
cols <- c("Phel_contig", "GO_ID", "GO_slim", "CC_MF_BP")
colnames(phelannot) <- cols
head(phelannot)
```

`join` degs and phelannot by "Phel_contig" column
```{r}
degsannotvs <- left_join(degstvs, phelannot, by = "Phel_contig")
head(degsannotvs)
```
389 degs, 10 columns

Write out table:
```{r}
write.table(degsannotvs, file = "../analyses/DESeq2/2015Phel/DEGlist_annot_4x4_tankexposed_V_sick.tab", row.names = FALSE, sep = '\t', quote = FALSE)
```
wrote out 20221204


